Slitasje og kjørespor

### 2023-01-16 DRAFT_slitasjeOgKjørespor.Rmd
doc: Fix headers

### Combine datasets

@@

The ANO data is point data, and the nature type data is multipolygon.

@@

The infrastructure index is explained
[here](https://brage.nina.no/nina-xmlui/handle/11250/2983607). I use it
as a kind of pressure or human influence index. It is likely to be a
good predictor for *slitasje*.

@@

```{r, eval=F}
infra <- stars::read_stars(paste0(path, "/NY_INFRA_IND.tif"))
```

@@

Transform to lat long dye to osm requirements

@@

```{r, eval=F}

@@

@@

I tweaked the thresholds so that most of the forest next to Trondheim
was i the second lowest class.

@@

I can scale the indicator for each polygon, or I can chose to aggregate
them first. If the scaled value is representative and precise at the
polygon level, then I could scale at that level. However, the combined
surveyed area is still so little of the total area of Norway, so that
only producing indicator values for these leaves the indicator without
much value for regional assessments. Later when we produce assessments
and indicator values at the regional level, we cannot simply do an area
weighting of the polygons in each region. This is because we don't want
to assume that the polygons are representative far outside of the mapped
area. But perhaps we can assume them to be representative inside
*homogenous ecological areas*. That's where the infrastructure index
comes in. Here's the plan:

@@

I will use the same reference levels/values for all of Norway:
```{r}
upper <- 0
lower <- 100
threshold <- 10
```

@@

I need to do a little trick and reverse the upper and lower reference values for the normalisation to work. This is a bug which can be fixed inside eaTools.
```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable."}
eaTools::ea_normalise(data = slitasje_data,
                      vector = "slitasje_percentage",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      plot=T,
                      reverse = T
                      )
```

This normalisation seems reasonable to me. I can save it as indicator values. I will assign the indicator value to the avergae year of the data.
```{r}
mean(slitasje_data$kartleggingsår)
```

@@

@@

## Prepare export

### Eksport file (final product)

<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

>>>>

##### Combine Naturtype data and ANO

@@

###### Points to polygons
The ANO data is point data, and the nature type data is multipolygon.
Because later we want to save this as a shape file, we cannot have a
mixed class type. I will therefore convert the points to polygons. I use
the area column to calculate a radii that gives that area.

```{r}
slitasje_data_points <- slitasje_data %>%
  mutate(g_type = st_geometry_type(.)) %>%
  filter(g_type =="POINT") %>%
  st_buffer(sqrt(slitasje_data$area/pi))
```

Checking no that the new polygons have the area corresponding to the
proportion of the point that was part of the same main ecosystem:

```{r new-area, fig.cap="Checking that the area of the new polygons fall in line with the proportion of each point which is part of the main ecosystem."}
slitasje_data_points$area2 <- st_area(slitasje_data_points)
plot(slitasje_data_points$area, slitasje_data_points$area2,
     xlab = "Target area",
     ylab = "Area of the new polygons")
```

The area calculation seems to have worked fine.

```{r}
slitasje_data_polygons <- slitasje_data %>%
  mutate(g_type = st_geometry_type(.)) %>%
  filter(g_type !="POINT")

slitasje_data <- bind_rows(slitasje_data_points, slitasje_data_polygons)

The new dataset only contains polygons.

@@

The infrastructure index is explained
[here](https://brage.nina.no/nina-xmlui/handle/11250/2983607). I use it
as a kind of pressure or human influence index. It is likely to be a
good predictor for *slitasje*, and I think to use it to define local reference area,
or homogeneous (impact) areas.

@@

```{r}
infra <- stars::read_stars(paste0(path, "/NY_INFRA_IND.tif"))
```

@@

Transform to lat long due to osm requirements

@@

```{r, eval=F}
path <- ifelse(dir == "C:", 
        "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks",
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks")

@@

Here I define a simplified categorical typology for the infrastructure
index using four classes.

@@

I tweaked the thresholds so that most of the forest next to Trondheim
was in the second lowest class. This looks pretty good to me.

@@

I can scale the indicator for each polygon, or I can chose to aggregate
them first. If the scaled value is representative and precise at the
polygon level, then I could scale at that level. However, the combined
surveyed area is still so little of the total area of Norway, so that
only producing indicator values for these leaves the indicator without
much value for regional assessments. Later when we produce assessments
and indicator values at the regional level, we cannot simply do an area
weighting of the polygons in each region. This is because we don't want
to assume that the polygons are representative far outside of the mapped
area. But perhaps we can assume them to be representative inside
*homogeneous ecological areas*. That's where the infrastructure index
comes in. Here's the plan:

@@

I will use the same reference levels/values for all of Norway:

```{r}
upper <- 0
lower <- 100
threshold <- 10
```

@@

I need to do a little trick and reverse the upper and lower reference
values for the normalisation to work. This is a bug which can be fixed
inside eaTools.

```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable."}
eaTools::ea_normalise(data = slitasje_data,
                      vector = "slitasje_percentage",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      plot=T,
                      reverse = T
                      )
```

This normalisation seems reasonable to me. I can save it as indicator
values. I will assign the indicator value to the average year of the
data.

```{r}
mean(slitasje_data$kartleggingsår)
```

@@

### Homogeneous impact areas

```{r, eval=F}
eaTools::ea_homogeneous_area(infra2)
```

@@

## Eksport files (final product)

According to the workflow specifications, I will export the indicator
map at its highest resolution, for use in various web interfaces etc.

But I also need to export a map of the indicator where we extrapolate the indicator values
to the homogeneous areas. This map should be interpreted with caution, since they seem
to provide explicit indicator values for areas where there is no data. This map is mainly for
aggregation purposes. I will mark this map with the prefix _homogeneous-areas_.

I will also export a map of the *homogeneous areas*, over which I will aggregate the indicator values.
This map might be useful in later indicator work.

For the finest resolution indicator map I will not include columns for
uncertainty (because this will only be estimated at the aggregated
level), or reference levels (because there is no spatial variation).

### Homogeneous areas
Export a map of the homogeneous areas that we aggregate the indicator
over:

```{r}
stars::write_stars(infra_trd_reclassed,
                   "data/homogeneous_areas/infrastructureIndex_fourClasses.tif")
```

### Sliatsje in wetlands
Prepare export
```{r}
exp_wetland <- select(slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",],
              origin,
              area,
              i_2020,
              SHAPE
              )
```

Write to file
```{r, eval=F}
st_write(exp_wetland, "indicators/indicatorMaps/slitasje/wetland_slitasje.shp",
         append = F)
```

### Sliatsje in wetlands, extrapolated to homegeneous impact areas.
